Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-15 paper/paper.tex
pass over section 6

@@

Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Coq,
using a recursion scheme in which termination is entirely enforced by its
typing.  This is a significant advance, since it avoids \emph{completely} the
need for termination proofs. Their work differs from ours in that they require
the functional extensionality axiom, and the use of impredicative \coq{Set}.
The authors justify well the use of impredicative \coq{Set} and its
compatibility with the functional extensionality axiom. In contrast, our
development remains entirely \emph{axiom-free}. Another key difference with our
approach is that they do not discuss what the resulting extracted code
\emph{looks like}.
% (that is, whether the extracted OCaml code resembles the
% natural formulation of the recursive function).
Through experiments, we found that their formalisation leads to
\ocaml{Obj.magic}, and code with a complex structure that may be hard to
understand or interface with other OCaml code. Due to the great benefit of
entirely avoiding termination proofs, it would be interesting to extend their
approach to improve code extraction.

@@

\subsection{Fusion}\label{sec:fusion}

Fusion in functional programming is a general technique for combining multiple
computations into a single pass to eliminate intermediate data structures and
improve efficiency. The fusion we have discussed in this paper relating to
program calculation is one form of fusion, but there are other approaches.

\emph{Deforestation}, introduced by Wadler~\cite{wadler88}, is a specific
approach to fusion that aims to remove intermediate trees (such as lists)
created between chained operations. A well-known method of deforestation is
\haskell{foldr/build} fusion~\cite{gill93}, which optimizes list-processing by
fusing list-producing functions (like \haskell{map} or \haskell{filter}) with
list-consuming ones (like \haskell{foldr}), avoiding the creation of temporary
lists.
%
Extending this idea, stream fusion generalizes the concept to handle other data
types, using a lazy stream abstraction to represent sequences and fuse
transformations like \haskell{map} and \haskell{fold}~\cite{coutts07}.
% Both approaches eliminate redundant structures and ensure computations happen in
% a single efficient pass over the data.

@@

In the context of Coq, Danvy explains the use of fold/unfold lemmas to mechanize
equational reasoning about recursive programs~\cite{DANVY_2022}. Fold-unfold
lemmas enable optimizations like fusion by allowing programs to be
systematically simplified, including eliminating intermediate structures.
Additionally, an approach to fusion called the worker/wrapper
transformation~\cite{gill09} was formalised and mechanised in
Agda~\cite{SCULTHORPE_HUTTON_2014}.

>>>>

\subsection{Fusion Optimisation}\label{sec:fusion}

Fusion in functional programming is a general technique for combining multiple
computations into a single pass to eliminate intermediate data structures or
repeated traversals, improving efficiency. The fusion we have discussed in this
paper relating to program calculation is one form of fusion, but there are other
approaches, and prior research has explored the mechanisation and verification
of fusion transformations.

Fusion has been applied to recursive tree traversals, in order to reduce the
number of times each section of the tree must be visited. Research has
demonstrated that this sort of fusion can be performed automatically by a
compiler~\cite{10.1145/3133900}, and the correctness of tree traversal fusion
has been verified mechanically in Coq~\cite{10.1145/3377555.3377884}.

Another related family transformations are worker/wrapper
transformations~\cite{gill09}, which have been used in the context of optimising
compilers, and have been mechanised in Agda~\cite{SCULTHORPE_HUTTON_2014}. This
technique has been applied to improving the performance of programs defined
using folds via fusion~\cite{hutton10}. Their work does not focus on the use of
an interactive proof assistant, but rather on the manual use of equational
reasoning to re-write programs and on the automatic transformation of functional
programs by a compiler.

@@

Abreu et al.~\cite{AbreuDHJMS23} encode divide-and-conquer computations in Coq,
using a recursion scheme in which termination is entirely enforced by its
typing. This is a significant advance, since it avoids \emph{completely} the
need for termination proofs. Their work differs from ours in that they require
the functional extensionality axiom, and the use of impredicative \coq{Set}. The
authors justify well the use of impredicative \coq{Set} and its compatibility
with the functional extensionality axiom. In contrast, our development remains
entirely \emph{axiom-free}, and we retain the ability to extract natural-looking
OCaml code.
% Another key difference with our
% approach is that they do not discuss what the resulting extracted code
% \emph{looks like}.
% (that is, whether the extracted OCaml code resembles the
% natural formulation of the recursive function).
Through experiments, we found that code extracted from their formalisation makes
heavy use of \ocaml{Obj.magic} and has a generally complex structure that may be
hard to understand and integrate with other OCaml code. Due to the great benefit
of entirely avoiding termination proofs, it would be interesting to extend their
approach to improve code extraction.

@@

% \subsection{Fusion}\label{sec:fusion}

% Fusion in functional programming is a general technique for combining multiple
% computations into a single pass to eliminate intermediate data structures and
% improve efficiency. The fusion we have discussed in this paper relating to
% program calculation is one form of fusion, but there are other approaches.

% \emph{Deforestation}, introduced by Wadler~\cite{wadler88}, is a specific
% approach to fusion that aims to remove intermediate trees (such as lists)
% created between chained operations. A well-known method of deforestation is
% \haskell{foldr/build} fusion~\cite{gill93}, which optimizes list-processing by
% fusing list-producing functions (like \haskell{map} or \haskell{filter}) with
% list-consuming ones (like \haskell{foldr}), avoiding the creation of temporary
% lists.
% %
% Extending this idea, stream fusion generalizes the concept to handle other data
% types, using a lazy stream abstraction to represent sequences and fuse
% transformations like \haskell{map} and \haskell{fold}~\cite{coutts07}.
% Both approaches eliminate redundant structures and ensure computations happen in
% a single efficient pass over the data.

@@

% In the context of Coq, Danvy explains the use of fold/unfold lemmas to mechanize
% equational reasoning about recursive programs~\cite{DANVY_2022}. Fold-unfold
% lemmas enable optimizations like fusion by allowing programs to be
% systematically simplified, including eliminating intermediate structures.
% Additionally, an approach to fusion called the worker/wrapper
% transformation~\cite{gill09} was formalised and mechanised in
% Agda~\cite{SCULTHORPE_HUTTON_2014}.
